### 2018-07-31 mlr_summary.tex
NN and PCA

@@

\section{Linear Regression}
One of the most basic algorithms for estimating a real valued function is linear regression.
Given a set of multivariate inputs $x_i \in \dom{R}^m$ and corresponding real valued outputs $y_i \in \dom{R}$ the following \enquote{loss} $L$ is to be minimized by a linear function $f_\beta:\dom{R}^m\to\dom{R}$.
\eqns{&
L(\beta) = \sum_{i=1}^n (\;y_i - f_\beta(x_i)\;)^2 
&&\\&
f_\beta(x) = x\T\beta \with\beta\in\dom{R}^m
&&}
This is the mean squared error and we want to minimize it using a linear function.
Lets call the matrix of transposed $x_i$'s $X$ and the vector of $y_i$'s $y$.
This allows for rewriting the loss as follows.
\eqnsnn{&
L(\beta) = (y-X\beta)\T(y-X\beta)
&&\\&
X = \mat{x_1\T\\\vdots\\x_n\T} \quad y = \mat{y_1\\\vdots\\y_n}
&&}
When minimizing we seek the roots of the derivative of the loss, by doing so we can derive the optimal solution for beta.
\eqnsnn{
\deriv{L(\beta)}{\beta} = 0 ~&\equivalent~ (y-X\beta)\T X + (y-X\beta)\T X = 2*(y-X\beta)\T X = 0
&&\\&
\equivalent (y-X\beta)\T X = X\T y - X\T X\beta = 0
&&\\&
\equivalent X\T y = X\T X\beta
&&\\&
\equivalent (X\T X)^{-1}X\T y = \beta
&&
}
\subsection{Non-Linear Features}
Sadly the very simple linear regression algorithm only allows for linear functions to be estimated. 
Fortunately we can make things non-linear by defining a feature function which maps an input $x$ to feature space like so:
\eqnsnn{
\phi(x) \in \dom{R}^l \with l\in\dom{N}
&&}
For real valued $x\in\dom{R}$ we can create the quadratic feature vector using the following.
\eqnsnn{
\phi(x) = \mat{1&x&x^2}\T
&&}
When using linear regression on these features we end up with a function that is quadratic in $x$ but linear in features $\phi(x)$.
\eqnsnn{&
X = \mat{\phi(x_1)\T\\\vdots\\\phi(x_n)\T}
&&\\&
f_\beta(x) = \phi(x)\T\beta = \phi(x)_1*\beta_1 + \phi(x)_2*\beta_2 + \phi(x)_3*\beta_3 = \beta_1 + x\beta_2 + x^2\beta_3
&&
}
Even more sophisticated features could be using gaussian functions at the datapoints which then amounts to a radial basis function.
\eqnsnn{&
\phi(x) = \mat{1 & \phi_1(x) & \cdots & \phi_N(x)}\T \with \phi_i(x) = e^{-\frac{1}{2}||x_i-x||_2^2}
&&}

\subsection{Regularization}
Maybe you may have noticed that the linear regression algorithm may fail in case that $(X\T X)$ is not invertible.
To ensure invertibility a tiny quantity can be added to or subtracted from the diagonal of the matrix like so: $(X\T X - \lambda I)$.
Interestingly, the optimal solution to the following loss has this exact form.
\eqnsnn{
L(\beta) = (y&-X\beta)\T(y-X\beta) + \lambda*\beta\T\beta
&&\\
\deriv{L(\beta)}{\beta} = 0 &\equivalent 2(y-X\beta)\T X + 2\lambda*\beta\T = 0
\equivalent (y-X\beta)\T X + \lambda*\beta\T = 0
&&\\
&\equivalent X\T y - X\T X\beta + \lambda*\beta = 0 
\equivalent X\T y = X\T X\beta - \lambda*\beta = (X\T X - \lambda I) \beta
&&\\
&\equivalent (X\T X - \lambda I)^{-1}X\T y = \beta
&&} 
The loss function can be viewed as a combination of penalties for $\beta$.
The first penalty is the mean squared error from before which is large when $\beta$ cannot reproduce the samples $y$ well.
The second penalty is for $\beta$ of long lenghts, which means that short $\beta$ with small coefficients are preferred.
These two penalties probably are contradictive and the optimal beta will be somewhere in between the individual minima for the penalties.
This kind of regularization using $\lambda\beta\T\beta$ is called $L^2$ regularization and is a special case of the general Tikhonov regularization formulating constraints on $\beta$ as $\beta\T\Gamma\T\Gamma\beta$ with Tikhonov matrix $\Gamma$.
$\Gamma\beta$ is a linear mapping which can be chosen to also formulate special constraints on beta like differences between coefficients, or projections to different vector spaces.

@@

\section{Logistic Regression}
We already discussed one classifier, the support vector machine, that is a binary classifier.
In this section we will take a look at a different approach to classification.
As usual we have multivariate $x_i \in \dom{R}^M$ as inputs which are labeled with a discrete variable $y_i \in \dom{Y}=\{Y_1,\dots,Y_l\}$ ($i = {1,\dots,N}$).
This time we want to know the probability of a class $y$ given that we observed some $x$ which is denoted by $P(y|x)$.

@@

Similar to the Boltzmann distribution $p(x) = e^{-E(x)}$ where $E(x)$ is an energy (or error), the probabilities are calculated as:
\eqnsnn{&
P(y|x) = \frac{e^{f(x,y)}}{Z} \mathtext{with normalization} Z = \sum_{y'\in\dom{Y}} e^{f(x,y')}
&&}

Now that we set up the basic framework, we need to define the discriminative function.
Lets formulate it as a linear function in features (similar to linear regression).
\eqnsnn{&
f(x,y) = \phi(x,y)\T\beta \with \phi(x,y) = \mat{\phi(x)[y=Y_1] \\ \phi(x)[y=Y_2] \\ \vdots \\ \phi(x)[y=Y_l]}
&&}
The $[cond]$ operator is $1$ if the condition is true, otherwise $0$. 
The objective of our model is to maximize the likelihood $p(y_i|x_i)$ for each $(x_i,y_i)$, but instead it could also maximize the log likelihood $\ln(p(y_i|x_i))$ since the logarithm is a monotonic increasing function.
Also, in machine learning we rather like to minimize stuff, so lets minimize the neg-log-likelihood by minimizing the following loss.
\eqnsnn{
L(\beta) &= -\sum_{i=1}^N \ln(\;p(y_i|x_i)\;) 
= - \sum_{i=1}^N\Big( \ln\big( e^{f(x_i,y_i)}\big) - \ln\big(\sum_{y' \in \dom{Y}} e^{f(x_i,y')}\big) \Big)
&&\\
&= - \sum_{i=1}^N\Big( \ln\big( e^{f(x_i,y_i)}\big)\Big) +\sum_{i=1}^N\Big( \ln\big(\sum_{y' \in \dom{Y}} e^{f(x_i,y')}\big) \Big) = -A(\beta) + B(\beta)
&&
}
Lets assume for simplicity that $\phi(x)$ is scalar, so that there is one $\beta_c$ corresponding to $\phi(x,y)_c$ with $c \in \{1,\dots,l\}$ indicating the class $Y_c$. 
\eqnsnn{
\deriv{L(\beta)}{\beta} &= -\deriv{A(\beta)}{\beta} + \deriv{B(\beta)}{\beta}
&&\\
\deriv{A(\beta)}{\beta} &= \sum_{i=1}^N \deriv{f(x_i,y_i)}{\beta} = \sum_{i=1}^N \phi(x_i,y_i)\T
&&\\
\deriv{B(\beta)}{\beta} &= \sum_{i=1}^N \deriv{}{\beta} \ln\big(\sum_{y' \in \dom{Y}} e^{f(x_i,y')}\big) 
&&\\
&= \sum_{i=1}^N \frac{1}{\sum_{y' \in \dom{Y}} e^{f(x_i,y')}} * \deriv{}{\beta} \sum_{y' \in \dom{Y}} e^{f(x_i,y')}
&&\\
&=\sum_{i=1}^N \frac{1}{Z} * \sum_{y' \in \dom{Y}} e^{f(x_i,y')}*\phi(x_i,y')\T
&&\\
&=\sum_{i=1}^N \sum_{y' \in \dom{Y}} 
\frac{e^{f(x_i,y')}}{Z} * \phi(x_i,y')\T
&&\\
&=\sum_{i=1}^N \sum_{y' \in \dom{Y}} p(y'|x_i) * \phi(x_i,y')\T
&&\\
\Big(\deriv{L(\beta)}{\beta}\Big)\T = \nabla_\beta L(\beta) &= \sum_{i=1}^N \Big( \sum_{y' \in \dom{Y}} p(y'|x_i) * \phi(x_i,y') \Big) - \phi(x_i,y_i)
&&}

@@

>>>>

\tableofcontents

@@

\section{Linear Regression}
One of the most basic algorithms for estimating a real valued function is linear regression.
Given a set of multivariate inputs $x_i \in \dom{R}^m$ and corresponding real valued outputs $y_i \in \dom{R}$ the following \enquote{loss} $L$ is to be minimized by a linear function $f_\beta:\dom{R}^m\to\dom{R}$.
\eqns{&
L(\beta) = \sum_{i=1}^n (\;y_i - f_\beta(x_i)\;)^2 
&&\\&
f_\beta(x) = x\T\beta \with\beta\in\dom{R}^m
&&}
This is the mean squared error and we want to minimize it using a linear function.
Lets call the matrix of transposed $x_i$'s $X$ and the vector of $y_i$'s $y$.
This allows for rewriting the loss as follows.
\eqnsnn{&
L(\beta) = (y-X\beta)\T(y-X\beta)
&&\\&
X = \mat{x_1\T\\\vdots\\x_n\T} \quad y = \mat{y_1\\\vdots\\y_n}
&&}
When minimizing we seek the roots of the derivative of the loss, by doing so we can derive the optimal solution for beta.
\eqnsnn{
\deriv{L(\beta)}{\beta} = 0 ~&\equivalent~ -(y-X\beta)\T X - (y-X\beta)\T X = -2*(y-X\beta)\T X = 0
&&\\&
\equivalent (y-X\beta)\T X = X\T y - X\T X\beta = 0
&&\\&
\equivalent X\T y = X\T X\beta
&&\\&
\equivalent (X\T X)^{-1}X\T y = \beta
&&
}
\subsection{Non-Linear Features}
Sadly the very simple linear regression algorithm only allows for linear functions to be estimated. 
Fortunately we can make things non-linear by defining a feature function which maps an input $x$ to feature space like so:
\eqnsnn{
\phi(x) \in \dom{R}^l \with l\in\dom{N}
&&}
For real valued $x\in\dom{R}$ we can create the quadratic feature vector using the following.
\eqnsnn{
\phi(x) = \mat{1&x&x^2}\T
&&}
When using linear regression on these features we end up with a function that is quadratic in $x$ but linear in features $\phi(x)$.
\eqnsnn{&
X = \mat{\phi(x_1)\T\\\vdots\\\phi(x_n)\T}
&&\\&
f_\beta(x) = \phi(x)\T\beta = \phi(x)_1*\beta_1 + \phi(x)_2*\beta_2 + \phi(x)_3*\beta_3 = \beta_1 + x\beta_2 + x^2\beta_3
&&
}
Even more sophisticated features could be using gaussian functions at the datapoints which then amounts to a radial basis function.
\eqnsnn{&
\phi(x) = \mat{1 & \phi_1(x) & \cdots & \phi_N(x)}\T \with \phi_i(x) = e^{-\frac{1}{2}||x_i-x||_2^2}
&&}

\subsection{Regularization}
Maybe you may have noticed that the linear regression algorithm may fail in case that $(X\T X)$ is not invertible.
To ensure invertibility a tiny quantity can be added to or subtracted from the diagonal of the matrix like so: $(X\T X - \lambda I)$.
Interestingly, the optimal solution to the following loss has this exact form.
\eqnsnn{
L(\beta) = (y&-X\beta)\T(y-X\beta) + \lambda*\beta\T\beta
&&\\
\deriv{L(\beta)}{\beta} = 0 &\equivalent -2(y-X\beta)\T X + 2\lambda*\beta\T = 0
&&\\
&\equivalent \lambda*\beta\T - (y-X\beta)\T X = 0
&&\\
&\equivalent \lambda*\beta - X\T y + X\T X\beta = 0 
&&\\
&\equivalent X\T y = X\T X\beta - \lambda*\beta = (X\T X - \lambda I) \beta
&&\\
&\equivalent (X\T X - \lambda I)^{-1}X\T y = \beta
&&} 
The loss function can be viewed as a combination of penalties for $\beta$.
The first penalty is the mean squared error from before which is large when $\beta$ cannot reproduce the samples $y$ well.
The second penalty is for $\beta$ of long lenghts, which means that short $\beta$ with small coefficients are preferred.
These two penalties probably are contradictive and the optimal beta will be somewhere in between the individual minima for the penalties.
This kind of regularization using $\lambda\beta\T\beta$ is called $L^2$ regularization and is a special case of the general Tikhonov regularization formulating constraints on $\beta$ as $\beta\T\Gamma\T\Gamma\beta$ with Tikhonov matrix $\Gamma$.
$\Gamma\beta$ is a linear mapping which can be chosen to also formulate special constraints on beta like differences between coefficients, or projections to different vector spaces.

@@

\section{Logistic Regression}
We already discussed one classifier, the support vector machine, that is a binary classifier.
In this section we will take a look at a different approach to classification.
As usual we have multivariate $x_i \in \dom{R}^M$ as inputs which are labeled with a discrete variable $y_i \in \dom{Y}=\{Y_1,\dots,Y_l\}$ ($i = {1,\dots,N}$).
This time we want to know the probability of a class $y$ given that we observed some $x$ which is denoted by $p(y|x)$.

@@

Similar to the Boltzmann distribution $p(x) = e^{-E(x)}$ where $E(x)$ is an energy (or error), the probabilities are calculated as:
\eqnsnn{&
p(y|x) = \frac{e^{f(x,y)}}{Z} \mathtext{with normalization} Z = \sum_{y'\in\dom{Y}} e^{f(x,y')}
&&}

Now that we set up the basic framework, we need to define the discriminative function.
Lets formulate it as a linear function in features (similar to linear regression).
\eqnsnn{&
f(x,y) = \phi(x,y)\T\beta \with \phi(x,y) = \mat{\phi(x)[y=Y_1] \\ \phi(x)[y=Y_2] \\ \vdots \\ \phi(x)[y=Y_l]}
&&}
The $[cond]$ operator is $1$ if the condition is true, otherwise $0$.

The objective of our model is to maximize the likelihood $p(y_i|x_i)$ for each $(x_i,y_i)$, but instead it could also maximize the log likelihood $\ln(p(y_i|x_i))$ since the logarithm is a monotonic increasing function.
Also, in machine learning we rather like to minimize stuff, so lets minimize the neg-log-likelihood by minimizing the following loss.
\eqnsnn{
L(\beta) &= -\sum_{i=1}^N \ln(\;p(y_i|x_i)\;) 
= - \sum_{i=1}^N\Big( \ln\big( e^{f(x_i,y_i)}\big) - \ln\big(\sum_{y' \in \dom{Y}} e^{f(x_i,y')}\big) \Big)
&&\\
&= - \sum_{i=1}^N\Big( \ln\big( e^{f(x_i,y_i)}\big)\Big) +\sum_{i=1}^N\Big( \ln\big(\sum_{y' \in \dom{Y}} e^{f(x_i,y')}\big) \Big) = -A(\beta) + B(\beta)
&&
}
The derivative of the loss is then calculated from the two separated sums $A(\beta)$ and $B(\beta)$:
\eqnsnn{
\deriv{L(\beta)}{\beta} &= -\deriv{A(\beta)}{\beta} + \deriv{B(\beta)}{\beta}
&&\\
\deriv{A(\beta)}{\beta} &= \sum_{i=1}^N \deriv{f(x_i,y_i)}{\beta} = \sum_{i=1}^N \phi(x_i,y_i)\T
&&\\
\deriv{B(\beta)}{\beta} &= \sum_{i=1}^N \deriv{}{\beta} \ln\big(\sum_{y' \in \dom{Y}} e^{f(x_i,y')}\big) 
&&\\
&= \sum_{i=1}^N \frac{1}{\sum_{y' \in \dom{Y}} e^{f(x_i,y')}} * \deriv{}{\beta} \sum_{y' \in \dom{Y}} e^{f(x_i,y')}
&&\\
&=\sum_{i=1}^N \frac{1}{Z} * \sum_{y' \in \dom{Y}} e^{f(x_i,y')}*\phi(x_i,y')\T
&&\\
&=\sum_{i=1}^N \sum_{y' \in \dom{Y}} 
\frac{e^{f(x_i,y')}}{Z} * \phi(x_i,y')\T
&&\\
&=\sum_{i=1}^N \sum_{y' \in \dom{Y}} p(y'|x_i) * \phi(x_i,y')\T
&&\\
\Big(\deriv{L(\beta)}{\beta}\Big)\T &= \nabla_\beta L(\beta) = \sum_{i=1}^N \Big( \sum_{y' \in \dom{Y}} p(y'|x_i) * \phi(x_i,y') \Big) - \phi(x_i,y_i)
&&}

@@

\section{Neural Networks}
Neural networks currently are the very hype in many fields of inteligent systems, probably because they are so flexible.
But lets not get carried away and start by defining a regression problem.
Given data $D=\{x_i,y_i\}_{i=1}^N$ with multivariate $x_i \in \dom{R}^M$ and scalar $y\in\dom{R}$ we want to estimate again as in linear regression $y=f(x)$.
For this, we model the function $f$ as follows:
\eqnsnn{&
f(x) = z_{L+1} = W_L*x_L \with x_l = \sigma(z_l) \;\;l\in\{2,\dots,L\} \mathtext{,} \sigma(z) = \frac{1}{1+e^{-z}} \mathtext{and} x_1 = x
&&}
In this equation we have $l$ matrices $W_1 \dots W_l$ with dimensions $W_1^{d(2)\times M}\; W_2^{d(3)\times d(2)}\dots W_l^{1\times d(l)}$. 
The sigmoid function $\sigma$ takes the role of the \enquote{activation function} which is an analogy to neurons in a brain which can propagate an action potential based on the magnitude of the signal.

Lets get more intuition on this from a simple example.
Suppose $x \in \dom{R}^3$ and $y \in \dom{R}^1$ and we have three matrices $W_1\;W_2\;W_3$.
To be able to input our $x$ into the neural net we have to choose the first dimension of $W_1$ to be $3$ and the last dimension of $W_3$ to be $1$ in order to map to $\dom{R}^1$. The other dimensions are left to be choosen freely. Lets choose $W_2^{10\times 10}$ which implies $W_1^{10 \times 3}$ and $W_3^{1 \times 10}$.
Our function is then:
\eqnsnn{&
f(x) = W_3\;\sigma\Big(W_2\;\sigma(W_1x)\Big)
&&}
This algorithm is called the forward propagation of $x$ through the neural network.
The unknowns in this equation are the entries of the matrices, the so called weights.
In order to optimize the weights we need to define a cost function $c(f,y)$ (a loss) for the prediction $f$.
In the literature, we can find many different cost functions that can be used, we'll stick to the well known mean squared error.
\eqnsnn{&
c(f,y) = (f-y)^2 \mathtext{so that} C_\text{MSE} = \frac1n \sum_{i=1}^N c(f(x_i),y_i)
&&\\&
\deriv{\;c(f,y)}{f} = 2(f-y)\T \mathtext{if $f$ and $y$ were vectors, in our case we can ommit the transpose}
&&}
Lets call $\deriv{\;c(f,y)}{f} = \delta_{L+1}$ as it is the derivative of the loss with repect to $z_{L+1}$ (in our case $\delta_4$ the derivative of $f(x) = z_4=W_3\;x_3$).
Due to the chain rule the derivatives for other layers are recursively calculated like this:
\eqnsnn{&
\delta_l = \deriv{c}{z_l} = \deriv{c}{z_{l+1}} * \deriv{z_{l+1}}{x_l}*\deriv{x_l}{z_l}
&&}
The derivative of the of the loss with respect to a specific weight of a matrix $W_{l,ij}$ is given by:
\eqnsnn{&
\deriv{c}{W_{l,ij}} = \deriv{c}{z_{l+1,i}}*\deriv{z_{l+1,i}}{W_{l,ij}}
&&}

To optimize the weights of the matrices, we feed the neural net all of our data samples $x_i$ and collect the losses $c(f(x_i),y_i)$.
We then backpropagate the losses through the net, which means we collect the derivatives of the losss for all weights and sum corresponding derivatives up forming a total gradient for a specific weight.
The weights are then updated using a gradient descend step which reads $W_{l,ij} = W_{l,ij} - \alpha \deriv{c}{W_{l,ij}}$ with learning rate $\alpha$ (probably $\alpha=0.1$).
And then we iterate forward and backward propagation.

\section{PCA}
Principle Component Analysis (PCA) is a method for dimensionality reduction.
This may be useful when you have very high dimensional data, but want to reduce the amount of data you want to process in order to learn a function.
In this case you may want to get a few of the \enquote{most important} parts of your highdimensional data point, or lets say a low dimensional datapoint which encodes the most important aspects of yout high dimensional point.

A way to do that, is to find a projection of our data $x_i \in \dom R^M$ onto $z_i \in \dom R^P$ where $P \leq M$.
Lets look at the projection from the other side, lets say we transform $z$ to $x$ using the following, where $V_P$ is the projection matrix and $\mu$ is a vector accounting for a translation in $\dom R^M$.
\eqnsnn{&
x \approx V_P z + \mu
&&\\&
err = \sum_{i=1}^N || V_P z_i + \mu - x_i ||^2
&&}
The error $err$ measures the mean squared error of the transformation of $z$ to $x$.
Using this, we can derive the optimal transformation of $x$ to $z$ when minimizing with respect to $z$.
\eqnsnn{
\hat{z}_{1:N} &= \argmin_{z_{1:N}} \sum_{i=1}^N || V_P z_i + \mu - x_i ||^2
&&\\
0 &= \deriv{}{z_i}\sum_{i=1}^N || V_P z_i + \mu - x_i ||^2 = \deriv{}{z_i} (V_P z_i + \mu - x_i)\T(V_P z_i + \mu - x_i)
&&\\
 &= 2*(V_P z_i + \mu - x_i)\T V_P
&&\\
\equivalent 0 &= V_P\T V_P z_i + V_P\T(\mu-x_i)
&&}
If we constrain $V_p$ to be orthonormal (composed of orthonormal vectors), then $V_P\T V_P = I$ and this simplifies to the follownig:
\eqnsnn{&
z_i = V_P\T(x_i-\mu)
&&}
Doing the same for $\mu$ yields the following:
\eqnsnn{
\hat{\mu} &= \argmin_{\mu} \sum_{i=1}^N || V_P z_i + \mu - x_i ||^2
&&\\
0 &= \deriv{}{\mu} \sum_{i=1}^N (V_P z_i + \mu - x_i)\T(V_P z_i + \mu - x_i)
 = 2* \sum_{i=1}^N (V_P z_i + \mu - x_i)\T 
&&\\
\equivalent 0 &= N*\mu*\sum_{i=1}^N (V_P z_i - x_i)
&&\\
\equivalent \mu &= \frac1N\sum_{i=1}^N V_Pz_i - \frac1N\sum_{i=1}^N x_i
&&}
So $\mu$ is the difference of the expeted value of the projected $z_i$ and the expected value $x_i$.
Lets assume that $E[z] = 0$ because that would be nice, and we don't care about the expected value of $z$ as it is unrelated to what we try to achieve. 
Due to the linearity of the expected value the projection of $z$ also has $0$ as expected value.
\eqnsnn{&
\mu = \frac1N\sum_{i=1}^N x_i = E[x]
&&}

So the projection $z_i = V_P\T(x_i-\mu)$ is the projecting the \enquote{centered} $x_i$.
For the derivation of $V_P$ lets assume that $x_i$ are already centered ($E(x)=0$).
What we want to minimize is the projection error, which we can estimate by back projecting the projection of $x_i$ and comparing it with the initial $x_i$.
\eqnsnn {&
\hat{V}_P = \argmin_{V_P} \sum_{i=1}^N ||x_i- V_PV_P\T x_i||^2 
&&}
